---
title: "An Introduction to the bigmemoryExtras Package"
author: "Peter M. Haverty"
date: "`r format(Sys.Date())`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
knit: rmarkdown::render
---

<!--
%\VignetteIndexEntry{bigmemoryExtras}
%\VignetteDepends{}
%\VignetteKeywords{bigmemoryExtras}
%\VignetteEngine{rmarkdown::render}
-->

```{r style, echo = FALSE, results='hide'}
BiocStyle::markdown()
```

# Introduction
This package defines a "BigMatrix" ReferenceClass which adds safety and convenience features to the filebacked.big.matrix class from the
bigmemory package. BigMatrix protects against segfaults by monitoring
and gracefully restoring the connection to on-disk data. We provide utilities for using BigMatrix-derived
classes as assayData matrices within the Biobase package's eSet family
of classes. BigMatrix provides some optimizations related to attaching
to, and indexing into, file-backed matrices with
dimnames. Additionally, the package provides a "BigMatrixFactor"
class, a file-backed matrix with factor properties.

```{r introcode}
library(bigmemoryExtras)
data.file = file.path(tempdir(),"bigmat","ds")
x = matrix(1:9,ncol=3,dimnames=list(letters[1:3],LETTERS[1:3]))
ds = BigMatrix(x,data.file)
ds[,1] = 3:1
ds[,1]
```

# Re-attaching to on-disk data as necessary
When a big.matrix object is attached to it's on-disk data, an external
pointer is used to connect the R object to a C++ data structure. When
a big.matrix object is not attached, like when it is loaded from an RData
file, this pointer is nil. Any access to this nil pointer will crash
R. The bigmemoryExtras package provides a BigMatrix class that prevents such
a crash by
controlling access to the external pointer. Additionally, BigMatrix
objects remember the location of their on-disk components and
automatically re-attach themselves as necessary.

This kind of thing would be helpful if you, for example, chose to save
your new BigMatrix object to disk for later use.  You might save your
object using R's built in save or saveRDS functions.

```{r reattach}
ds$backingfile
saveRDS(ds,file=file.path(tempdir(),"foo.rds"))
new.ds = readRDS(file=file.path(tempdir(),"foo.rds"))
new.ds[1:2,2:3]
```

# S4 Style Access
The BigMatrix class uses R's Reference Class system. Any change to the
matrix portion of the data has on-disk side effects, so it seems
natural that any other changes to the object should have the same
behavior. In order to give BigMatrix the same API as a base matrix or
big.matrix class, certain S4-style methods are provided.
ReferenceClass objects are relatively new to R and unfamiliar to many
users, so you may want to review the ReferenceClasses help page.

```{r S4methods}
nrow(ds)
ds$nrow()
ncol(ds)
ds$ncol()
dim(ds)
ds$dim()
dimnames(ds)
ds$dimnames()
length(ds)
ds$length()
```

# BigMatrixFactor
The bigmemoryExtras package adds a ``BigMatrixFactor'' class to provide a
means to store large matrices of characters. On the file system, these
are stored as the C type char or int (8 or 32 bits), depending on the number of
levels in the factor. Subsetting a BigMatrixFactor returns a factor. If more than one column is returned, the returned object is a factor matrix.

```{r factorcode}
data.file = file.path(tempdir(),"bigmat","fs")
x = matrix( c(rep("AA",5),rep("BB",4)) ,ncol=3,dimnames=list(letters[1:3],LETTERS[1:3]))
fs = BigMatrixFactor(x,data.file,levels=c("AA","BB"))
fs[,]
as(fs,"matrix")
fs$levels
levels(fs)
fs[, 2]
fs[, 2:3]
```

# Use with GenomicRanges and SummarizedExperiment-derived Classes

Either class can be used as an assay in the assays slot
of the GenomicRanges SummarizedExperiment-derived classes. We provide utility
functions to deal with relocated BigMatrix in such a container.

```{r se}
library(GenomicRanges)
se = SummarizedExperiment(assays=list(a=ds, b=fs))
assays(se)[["a"]]
new.dir = file.path(tempdir(),"newbigmat")
dir.create(new.dir,showWarnings=FALSE)
file.copy(ds$backingfile, new.dir)
file.copy(fs$backingfile, new.dir)
assays(se) = updateBackingfiles(assays(se), new.dir)
assays(se)[["b"]]$backingfile
```

# Use with Biobase and eSet-derived Classes
We are phasing out Biobase and eSet in favor of GenomicRanges and
SummarizedExperiment, but in the mean time ...

Either class can be used as an assayDataElement in the assayData slot
of the familiar BioConductor eSet-derived classes. We provide utility
functions to deal with relocated BigMatrix files.

```{r eset}
library(Biobase)
eset = ExpressionSet()
data.file = file.path(tempdir(),"bigmat","ds")
x = matrix(1:9,ncol=3,dimnames=list(letters[1:3],LETTERS[1:3]))
ds = BigMatrix(x,data.file)
assayDataElement(eset,"exprs") = ds
exprs(eset)[1:2,2:3]

new.dir = file.path(tempdir(),"newbigmat")
dir.create(new.dir,showWarnings=FALSE)
file.copy(ds$backingfile, new.dir)
assayData(eset) = updateBackingfiles(assayData(eset), new.dir)
assayDataElement(eset,"exprs")$backingfile
```
